### 2012-09-07 joint action and knowing others minds/conf budapest ceu talk cognitive 2012-09/interacting mindreaders ceu talk.tex
moved folders

\def \ititle {Interacting Mindreaders}
\def \isubtitle {}
\def \iauthor {Stephen A. Butterfill}
\def \iemail{s.butterfill@warwick.ac.uk}
%\def \iauthor {}
%\def \iemail{}
\date{}

@@

Mindreading is
the process of identifying thoughts and actions on the basis of bodily movements,
somewhat as reading is the process of identifying propositions on the basis of inscriptions \citep[p.\ 4]{Apperly:2010kx}.
Contrast
a mindreader who is, or appears to be, capable of interacting with her targets 
and
a mindreader who can manifestly only observe.
Is it possible that the interacting mindreader is in a position to know things which she would be unable to know if she were unable to interact with her targets?
Our aim in this paper is to argue that the answer is a qualified `yes'.

The question is about the evidential basis of mindreading, 
not about what sorts of mechanisms are involved.
While philosophers have engaged with questions about mechanisms (such as whether mindreading involves a process of simulation or of theorizing or some combination of the two),
comparatively little effort has recently been devoted to issues about what evidence could ground mindreading.

Our question is part of a broader question,
What is the evidential basis for ascriptions of thought and action and how does the evidence support the ascriptions?
%\footnote{
%See e.g.\ \citet[pp.\ 126-7]{Davidson:1973jx}.
%}
%To stress that this question is not directly about about \emph{how} anyone actually ascribes thoughts or identifies actions% 
%% nor directly about the mechanisms, processes or representations involved in mindreading
%, Davidson 
%sometimes formulates the question by asking what someone \emph{could} know that \emph{would} put them in a position to identify another's thoughts and actions.%
%\footnote{
%See, e.g., \citet[p.\ 126]{Davidson:1973jx}.
%While what follows draws  on Davidson's questions and insights,
%our aims here are more modest than his.
%For here we are concerned only with a  fraction of the problem of ascribing mental states and meanings,
%and we are not concerned with larger claims about mind, meaning and truth 
%(on which see 
%\citet{Davidson:1990du} and 
%\citet{lepore_donald_2005}).
%}
The most sustained attempts to answer this question,
Davidson's (\citeyear{Davidson:1984wh}; \citeyear{Davidson:1990du}), Lewis' (\citeyear{lewis:1974ri}) and Dennett's (\citeyear{Dennett:1987sf}),
do not exploit the possibility of interaction.
The evidence and principles they consider are available to manifestly passive observers.
So on their theories,
a purely passive mindreader observing from behind a one-way mirror
is on a par with
a mindreader who, actually or apparently, could interact with those she seeks to interpret.
The two are on a par in this sense:
in principle the same evidence could be available to each, and each can exploit the routes to knowledge in moving from evidence to ascriptions of thought and action.
Of course these theories are all compatible with the idea that interaction might be useful for mindreading in practice.  
But on these theories interaction makes no difference to what can in principle be known.
In this paper we aim to show
that mindreaders actually or apparently capable of interacting with their targets are at an advantage not only in practice but also in theory.
Their ascriptions could exploit routes to knowledge which would be unavailable if they were entirely passive observers.

Why suppose, in advance of considering the details, that interacting mindreaders might know more?
A mindreader's target is often also a mindreader and may sometimes reciprocate by taking the mindreader as a target for mindreading.
%Where this happens we have mindreaders mindreading and being mindread by mindreaders.
%The mindreader is her target's target for mindreading.
It ought to be possible, in mindreading, to make use of this reciprocity.
%of the fact that the mindreader's target reciprocates by taking the mindreader to be a target for mindreading.
But how could such reciprocity facilitate mindreading?
If we assume the mindreader is merely observing her target,
that there is manifestly no potential for interaction,
then it seems that any way of exploiting reciprocity would involve higher-order ascriptions. 
The mindreader would ascribe to her target beliefs (say) about the mindreader's own beliefs and other mental states.
And if her target reciprocates, she might escalate by ascribing to the target beliefs about her own beliefs about the target's beliefs about her beliefs.
While this might be useful in some situations, 
the nesting this approach requires quickly becomes dauntingly complex.
And the basic intuition about reciprocal mindreading goes unsatisfied.
Reciprocal mindreading should sometimes result in 
something like a meeting of minds
rather than
an escalation of higher-order ascriptions.
Perhaps fully exploiting reciprocity in mindreading
requires the mindreaders to be in a position to interact with each other;
perhaps in some cases 
being or appearing poised to interact
can somehow enable mindreaders to exploit reciprocity without first having to ascribe higher-order mental states.
This is the hunch we develop in what follows.

Even if the hunch turns out to be right, why investigate it?
One reason is that the investigation will enable us to revise and extend existing accounts of the evidential basis of mindreading  in ways that make them more accurate and comprehensive.
Another motive concerns the emergence, in evolution or in development, of mindreading.
Several researchers have offered quite general conjectures about how interaction might explain the emergence of sophisticated forms of cognition.
One view is that needs to interact with others have driven and shaped some aspects of cognition.%
\footnote{
E.g.\ \citet[p.\ 103]{Knoblich:2006bn} suggest that 
`functions traditionally considered hallmarks of individual cognition originated through the need to interact with others'
and that
`perception, action, and cognition are grounded in social interaction.'
}
We shall not speak to this view here.
Another view is that abilities to interact with others
may have fostered the emergence in evolution---or, on a different view, in development---%
of sophisticated forms of cognition including social cognition and, in particular, mindreading.%
\footnote{
\citet[p.\ 1]{Moll:2007gu}
argue for the `Vygotskian Intelligence Hypothesis' according to which `the unique aspects of human cognition ... were driven by, or even constituted by, social co-operation.'
See also
	\citet{Hughes:2004zj},
	\citet{Hughes:2006fu},
	\citet{Tomasello:2007gl} and
	\citet{tomasello:2008origins}.
}
Studying how interaction broadens the evidential basis of mindreading will eventually point to one way of filling in some details on just how interaction might facilitate the emergence of sophisticated forms of mindreading.

Our claim needs to be qualified in several ways.
First, we should highlight something already explicit but not yet emphasised:
the claim applies not just to 
 mindreaders who actually could interact with their targets
but also to
 mindreaders who appear to their targets to be potential interaction partners (even if they are not).
Trading the risk of a minor misunderstanding for concision we use the term `interacting mindreader' to refer to both groups;
similarly, references to mindreaders who `only observe' must be understood as excluding those who appear to their targets to be interaction partners.
Only cases involving the first group---mindreaders who actually could interact---are likely to be of wider interest.
%Cases involving misleading appearances might be regarded as artefacts of the model.
Any application of our claim to understanding evolution or development is bound to focus on cases where mindreaders actually could interact with their targets.

A second qualification concerns the scope of mindreading.
In most discussions of mindreading, the focus is on  ascription of beliefs and other mental states to individuals.
But in what follows we shall focus on the ascription of goals to actions.
Some might claim that goal ascription is not mindreading, perhaps because 
identifying relations between actions and the outcomes to which they are directed
does not necessarily involve ascribing mental states.
Our view is consistent with this claim (and with its negation).
\label{goal_ascription_verify}
Evidence for ascriptions of belief and of other mental states often includes the occurrence (and non-occurrence) of goal-directed actions.
Identifying these actions typically requires goal ascription.
In some cases, then, 
evidence for the ascription of a goal
will indirectly support the ascription of a belief or other mental state---%
the evidence will support  a goal ascription which will in turn support the ascription of a belief (say) to the agent of the action.
Accordingly, even those who deny that goal ascription is mindreading should agree that sometimes evidence for goal ascription is indirectly evidence for ascription of belief and other mental states.
So whether or not goal ascription is deemed to be mindreading,
an individual's access to evidence for mindreading  depends in part on her access to evidence for goal ascription.
Our plan is to show that interacting mindreaders could exploit routes to knowledge of the goals of others' actions which are not available to mere observers.
In doing this we will be showing that
mindreaders poised to interact with others 
could know things about their minds 
which they might not otherwise be in a position to know.

%Because the leading theories have supposed that being able to interact makes no difference,
%they have treated the evidential basis for mindreading as more narrow than it truly is.
%This matters
%because broadening the evidential basis will enable us to give a more accurate and comprehensive theory of mindreading.
%And, as we shall argue, this in turn matters because it may bear on the possibility of explaining the emergence, in evolution or in development, of mindreading.

Developing these ideas requires us to fill in some background on goal ascription and its limits, as well as on notions of goal-directed interaction.
Readers impatient to get to the central idea might skip to section \vref{sec:your_goal_is_my_goal}.

\section{Goal ascription}
Purposive action is action directed to the realisation of one or more outcomes.
Goal ascription is the process of identifying to which outcomes others' purposive actions are directed.
To illustrate, suppose that
Hannah kicks a ball thereby both preventing her sisters from scoring and also breaking a window.
Asked about the episode,
Hannah might protest, truthfully, that the goal of her action was not to break the window but only to reverse the others' advance.
As this illustrates,
among the actual and possible outcomes of an action,
only some are outcomes to which the action is directed.
Goal ascription is the process of identifying 
 those outcomes.
%, among an action's actual and possible outcomes,
%those to which it is directed.
%Our immediate question is therefore, 
%What evidence could support hypotheses about the outcomes to which actions are directed, 
%and how would the evidence support the hypotheses?

We focus on goal ascription 
partly because this simplifies our argument,
but mainly because goal ascription is widely thought to be among the very earliest components of mindreading to emerge (or, if goal ascription is not mindreading, then it is a late precursor).%
\footnote{
See, for example,
\citet{Gergely:1995sq} and 
\citet{Woodward:1998dm}.
See also
\citet[p.\ 111, Box 1]{Baillargeon:gx}
on two subsystems
and 
\citet{Povinelli:2001jf} on `behavioural regularities' whose specification sometimes appears to involve goal-directed action.
}
By showing that
  interacting mindreaders may have access to 
  evidence for goal ascriptions 
  which is unavailable to those who merely observe,
we will eventually be able to indicate ways in which interaction could facilitate the emergence, in evolution or development, of more sophisticated forms of mindreading.%
\footnote{
There are various ways of understanding the idea that mindreading may come in several forms, some less conceptually or cognitively demanding than others.
For a range of views, see
 \citet{Apperly:2009ju},
 \citet{Call:2005qe},
 \citet{Doherty:2006wz},
 \citet{ONeill:2005ff}
 and
 \citet{Wellman:2001if}.
}

Because goal ascription has received little attention,
in this section
we shall 
briefly review some potential benefits of being able to identify goals
and then 
consider what sort of evidence might support goal ascription.

It is a familiar idea that goal ascription enables one to learn from others' successes.
For example,
if you know or can guess that another agent's actions are directed to opening a nut,
you may then be in a position to infer that the unfamiliar pattern of actions she is performing constitute a means to open nuts.
%\citep{Horner:2005pj}
%In a different case, knowing the goal of another agent's actions may enable you to discover a new use for a tool.
A slightly less familiar idea is that goal ascription enables one to learn from others' failures as well as their successes.
For example, suppose that while you are searching for some peanuts 
another agent attempts but fails to reach for a closed container.
In some circumstances,
if you know that the goal of the agent's action was to obtain the peanuts
then you now have evidence as to where they might be.\footnote{
\citet{hare_chimpanzees_2004} exploit this fact in testing chimpanzees' abilities to ascribe goals.
}
This is one illustration of how goal ascription could in principle enable us to learn from others' failures 
(\citealp{Want:2001hp} offer another).

Goal ascription enables one
to predict and manipulate others' actions.
If you know that an agent is engaged in a sequence of actions whose eventual goal involves retrieving some object,
you may be able to predict that the agent will go to where the object is at some point.\footnote{
For an application see \citet{Hare:2001ph}.
}
Equally, in some cases knowing this much about the goal of another's actions many enable you to assist them,
either 
by retrieving the object for them 
or else
by revealing the object's location to them.%
\footnote{
For a paradigm involving the former see \citet{warneken:2007sa};
on the latter, see \citet{Liszkowski:2008al}.
}

Goal ascription is also instrumental for ascribing propositional attitudes.
We have already mentioned 
   (\vpageref*{goal_ascription_verify})
that such ascriptions often involve 
confirming predictions about actions 
which in turn typically requires
identifying the goals of those actions.
In addition,
knowing which outcomes an action is directed to may constrain hypotheses about what an agent intends 
as well as
potentially providing information concerning what the agent knows, believes or desires.
For example,
if we know that the goal of an agent's action is to retrieve some peanuts,
and if we also know where all the peanuts are,
we may be able to infer that she does not know where the peanuts are,
or that she falsely believes that some of the peanuts are over there.%
\footnote{
\citet{Wimmer:1998kx} exploit this possibility in testing children's abilities to ascribe false beliefs.
}
(Of course this can also work the other way:
information about an agent's beliefs or other mental states may support conclusions about her goals.
Belief- and goal-ascriptions are mutually constraining.)

Now that we have reviewed some of the benefits goal ascription can bring,
what evidence could support ascriptions of goals to actions?
Consider the claim that 
knowledgeably identifying the goals to which actions are directed
depends on 
knowledge of the agents' intentions
which in turn depends 
on knowledge of their beliefs, desires and other mental states.
If this claim were true,
it would make no sense to discuss the evidential basis of goal ascription except in discussing the evidential basis of mindreading more generally.
However,
the falsity of this claim  is presupposed  in both developmental and comparative research on goal ascription.%
\footnote{ 
Compare \citet{Gergely:1995sq},
	\citet{Woodward:1998dm} and
	\citet{Penn:2007ey}
among many others.
}
And we know of no compelling reason for taking this claim to be true.
After all,
it is consistent with rejecting the claim to recognize that
 ascriptions of goals to actions
 constrain, and are constrained by,
 ascriptions of intentions and other mental states.
The existence of such constraints does not show that one could 
not know something about the goals of actions 
while  knowing nothing about the agents' mental states.
Further,
there do seem to be situations where knowledge of agents' goals does not require any knowledge of their mental states.
For instance,
suppose that two people are sitting opposite each other at a low table
 which is 
sparsely populated with objects.
The objects are all out in the open; manifestly, both can clearly see them.
If one person reaches to grasp one of these objects (the duck, say), 
must the other ascribe beliefs or other mental states in order to knowledgeably identify the goal of her action as that of grasping the duck?
On the face of it, she need not.  
Even if she had no ability to ascribe mental states, it seems she might nevertheless be in a position to identify the goal of the other's action.

While not decisive,
these considerations are perhaps sufficient to motivate 
exploring what evidence might support goal ascription.
According to 
what Csibra and Gergely call `the principle of rational action',
%
\begin{quote}
`an action can be explained by a goal state if, and only if, it is seen as the most justifiable action towards that goal state that is available within the constraints of reality.'%
\footnote{
\citet[p.\ 255]{Csibra:1998cx}; cf.\ \citet{Csibra:2003jv}.
A related but different `principle of efficiency' has been formulated by \citet[p.\ 1061]{Southgate:2008el}:
`goal attribution requires that agents expend the least possible amount of energy within their motor constraints to achieve a certain end.'
}
\end{quote}
%
Taking this idea as a rough starting point,
we propose that
%goal ascription could sometimes involve 
%identifying which potentially desirable outcomes 
%an observed action is the best available means to achieving.
%More carefully, our proposal is that 
these facts:
%
\begin{enumerate}
%
\item action $a$ is directed to some goal;
%
\item actions of $a$'s type are normally capable of being means of realising outcomes of $G$'s type in situations with the salient (to any concerned) features of this situation;
% 
\item no alternative type of action is both 
typically available to agents of this type 
and also 
such that actions of this type would be normally be significantly better\footnotemark \ means of realising outcome $G$ in situations with the salient features of this situation;
\footnotetext{
An action of type $a'$ is a better means of realising outcome $G$ in a given situation than an action of type $a$ if, for instance, actions of type $a'$ normally involve less effort than actions of type $a$ 
in situations with the salient features of this situation 
and everything else is equal; 
or if, for example, actions of type $a'$ are normally more likely to realise outcome $G$ than actions of type $a$
in situations with the salient features of this situation 
and everything else is equal.
}
%
\item the occurrence of outcome $G$ is typically desirable for agents of this type;
%
\end{enumerate}
%
and
%
\begin{enumerate}[resume]
\item there is no other outcome, $G'$, 
the occurrence of which would be at least comparably desirable for agents of this type 
and where (2) and (3) both hold of $G'$ and $a$
%
%where
%	the occurrence of $G'$ would be at least comparably desirable for agents of this type 
%%where
%%	$G$'s occurrence is not a means to $G'$'s nor conversely,
%and
%where  
%	(2) and (3) both hold of $G'$ and $a$
\end{enumerate}
%
may jointly constitute defeasible evidence for the conclusion that:
%
\begin{enumerate}[resume]
\item $G$ is a goal to which action $a$ is directed.
\end{enumerate}
%
We suggest that the above inference, from (1)-(5) to (6), is a route to knowledge of the goals of actions in this sense:
in some cases
it would be possible to know the premises without already knowing the conclusion;
and, 
in some of those cases,
knowing the premises could put one in a position to know the conclusion.%
\footnote{
Knowledge of the conclusion may not require explicit knowledge of 
	what is salient to others,
	what is desirable to them,
	or what makes actions better for them.
It is arguably sufficient that
the individual ascribing a goal 
is entitled to 
rely on being sufficiently similar to the target of her ascription 
with respect to 
these things.
%	what makes one action a better or worse means of realising an outcome than another action for her,  
%	what is desirable to her,
%	and what is salient to her.
}

Why accept this?
Suppose that 
	(1)-(5) are facts 
and that
	$G$ is not a goal to which action $a$ is directed.
Then from (1) we know that $a$ has a goal, call it $G'$.
And from (5) we can infer that either 
(i) 
the action, $a$, was not of a type that is normally capable 
of being a means to the outcome to which it was directed,
or
(ii) 
a significantly better (for instance, a more reliable or less effortful) means of achieving $G'$ is typically available to agents of this type,
or
(iii)
$G'$'s occurrence would be significantly less desirable than $G$'s occurrence for agents of this type.
So the agent's action fell short of being appropriate or optimal in some way,
or the agent has atypical preferences or capabilities,
or the situation is not normal.
While any of these is possible---%
and perhaps in some cases even to be expected---%
there are also circumstances in which it is reasonable to suppose that none obtains,
that the situation will be normal in relevant respects,
that agents will have typical preferences and capabilities 
and that they will act in ways that are appropriate and (in the limited sense in play here) optimal.
In such circumstances, the above inference can serve as a route to knowledge of the goals of actions.
The existence of this route to knowledge shows that goal ascription doesn't invariably depend on mental state ascription.

Note that we are not suggesting that the above inference captures the only route to knowledge of the goals of actions,
or that it provides anything like a comprehensive theory of the evidential basis of goal ascription.
Clearly it does not.
The inference has only limited applications.
For example, it cannot be used when an action is the best available means of achieving two or more comparably desirable outcomes. 
Fortunately the argument that follows does not depend on having a comprehensive theory of evidence for goal ascription.
The point of this brief discussion was only to 
consider the sort of evidence that can bear on goal ascription
and to 
show that 
even those incapable of ascribing mental states like belief 
might nevertheless have evidence sufficient for goal ascription.

Our narrow aim in what follows is to show that some routes to knowledge of the goals of actions 
are available only to interacting mindreaders
and not to those who merely observe.
Having in this section introduced goal ascription as a topic,
the next step is to identify an obstacle to acquiring knowledge of the goals of actions,
one which we will eventually show can be avoided where mindreaders can (or appear able to) interact with their targets.

\section{The problem of opaque means}
\label{sec:opaque_means}

While we lack a detailed theory of the evidential basis of goal ascription,
it is certain that the evidence for goal ascription sometimes includes considerations about which ends actions are means to.
Suppose an observer faces an action but cannot identify ends to which it could be a means.
This may prevent her from recognizing the action's goal%
\footnote{
It is possible that some actions have more than one goal.
To reduce parenthetical qualifications we shall write as if actions had only one goal.  
All of our key claims and arguments are consistent with the possibility of actions with more than one goal.
} 
 by depriving her of evidence.
To illustrate, contrast two cases of tool use.
In one case, someone uses a reamer to  juice  a lime; in the other, someone else scores shag with a lame to prevent a loaf from cracking.
Without communication, repetition or convention,
an observer familiar with reamers but not lames 
may be able to identify the goal of the first action only.
As this illustrates, ignorance about to which ends actions are means can be an obstacle to goal ascription.
Call this the problem of opaque means.

We are not suggesting that no observer could ever identify the goal of any action she fails to recognise as a means to achieving that goal.
Of course opaque means are not in every case an insurmountable obstacle to goal ascription, and
they may only rarely be a problem for human adults with sophisticated social skills.
Our point is neither novel nor surprising:
opaque means sometimes  deprive mindreaders of evidence and so prevent goal ascription.
This is more likely to happen where  goal ascribers lack sophistication in mindreading, communication and culture.

Some of the most plausibly unique aspects of human cognition depend on our abilities to recognise the goals of novel behaviours involving tools, and of communicative gestures.
The problem of opaque means is likely to arise in both cases
if goal ascription is based  entirely on observation 
(so that the possibility of interaction is ignored)
and if the goal ascriber lacks sophisticated social skills.
We have just seen an illustration of how the problem of opaque means arises where tools are used to unfamiliar ends.
Relatedly, it is also likely to arise where actions involve multiple steps that do not form a familiar sequence, can occur in various orders and can be interspersed among other activities;
as in preparing spirit from grain, for example.
%\footnote{
%*cut because 'plausibly unique aspects of human cognition'*
%one example is food preparation in some groups of mountain gorillas
%\citep[p.\ 531]{Byrne:2003wx}.
%*This is slightly tricky because Byrne wants a pure behaviour reading account with no understanding of intention.  
%Stress that this is consistent with supposing that goals are ascribed.
%}

>>>>

\def \ititle {Interacting Mindreaders:}
\def \isubtitle {CEU Talk}
\def \iauthor {Stephen A. Butterfill}
\def \iemail{s.butterfill@warwick.ac.uk}
%\def \iauthor {}
%\def \iemail{}
\date{}

@@

Thank you for inviting me.  I’m visiting until Christmas to immerse myself in your work here, with the aim continuing one collaboration and perhaps starting new collaborative work.

My work is mostly on mindreading and joint action.
Here I'm aiming to bring the two together by discussing mindreading by the agents of a joint action.
The question I aim to answer is  whether interacting mindreaders could  be in a position to know things which they would be unable to know if they were manifestly passive observers.

But before I get to the question,
 let me step back to explain the wider issues that motivate it.

\section{slide: Vygotskian Intelligence Hypothesis}
Moll and Tomasello say ...

Coming at things from a very different starting point, 
Sebanz and Knoblich make a different but related claim.  
They say ...

Faced with these bold conjectures, it is natural to ask,
How could that work?
How could social interaction drive or ground sophisticated forms of cognition such as those involved in communication by language or sophisticated forms of mindreading?

I think we might be able to make a small step towards understanding these conjectures by considering the role of  social interaction in simple forms of mindreading.

\section{slide: Question}
So this talk is based on a narrow question.
Contrast
a mindreader who is, or appears to be, capable of interacting with her targets 
and
a mindreader who can manifestly only observe.
Is it possible that the interacting mindreader is in a position to know things which she would be unable to know if she were unable to interact with her targets?

Most philosophers who have written about the evidential basis of mindreading have assumed that the answer is `no'.
Of course they would allow that interaction can facilitate mindreading in practice.
But their theories on the evidential basis of mindreading assume that interacting mindreaders have no advantage \emph{in principle} over mere observers.
So on their theories,
a purely passive mindreader observing from behind a one-way mirror
is on a par with
a mindreader who, actually or apparently, could interact with those she seeks to interpret.
The two are on a par in this sense:
in principle the same evidence could be available to each, and each can exploit the routes to knowledge in moving from evidence to ascriptions of thought and action.

In this talk I aim to show that mindreaders actually or apparently capable of interacting with their targets are at an advantage not only in practice but also in theory.
Their ascriptions could exploit routes to knowledge which would be unavailable if they were entirely passive observers.

\section{slide: Reciprocity}
Why suppose, in advance of considering the details, that interacting mindreaders might know more?
A mindreader's target is often also a mindreader and may sometimes reciprocate by taking the mindreader as a target for mindreading.
%Where this happens we have mindreaders mindreading and being mindread by mindreaders.
%The mindreader is her target's target for mindreading.
It ought to be possible, in mindreading, to make use of this reciprocity.
%of the fact that the mindreader's target reciprocates by taking the mindreader to be a target for mindreading.
But how could such reciprocity facilitate mindreading?
If we assume the mindreader is merely observing her target,
that there is manifestly no potential for interaction,
then it seems that any way of exploiting reciprocity would involve higher-order ascriptions. 
The mindreader would ascribe to her target beliefs (say) about the mindreader's own beliefs and other mental states.
And if her target reciprocates, she might escalate by ascribing to the target beliefs about her own beliefs about the target's beliefs about her beliefs.
While this might be useful in some situations, 
the nesting this approach requires quickly becomes dauntingly complex.
And the basic intuition about reciprocal mindreading goes unsatisfied.
Reciprocal mindreading should sometimes result in 
something like a meeting of minds
rather than
an escalation of higher-order ascriptions.
Perhaps fully exploiting reciprocity in mindreading
requires the mindreaders to be in a position to interact with each other;
perhaps in some cases 
being or appearing poised to interact
can somehow enable mindreaders to exploit reciprocity without first having to ascribe higher-order mental states.
This is the hunch we develop in what follows.

\section{slide: Goal Ascription}

In most discussions of mindreading, the focus is on  ascription of beliefs and other mental states to individuals.
But in what follows I want to focus on goal ascription.

Goal ascription is the process of identifying to which outcomes others' purposive actions are directed.
To illustrate, suppose that
Hannah kicks a ball thereby both preventing her sisters from scoring and also breaking a window.
Asked about the episode,
Hannah might protest, truthfully, that the goal of her action was not to break the window but only to reverse the others' advance.
As this illustrates,
among the actual and possible outcomes of an action,
only some are outcomes to which the action is directed.
Goal ascription is the process of identifying 
 those outcomes.

Some might claim that goal ascription is not mindreading, perhaps because 
identifying relations between actions and the outcomes to which they are directed
does not necessarily involve ascribing mental states.
I'm happy if you want to define mindreading so as to exclude mere goal ascription.

Still, everyone should agree that mindreading very often depends on goal ascriptions. 
After all,
evidence for ascriptions of belief and of other mental states often includes the occurrence (and non-occurrence) of goal-directed actions.
Identifying these actions typically requires goal ascription.
In some cases, then, 
evidence for the ascription of a goal
will indirectly support the ascription of a belief or other mental state---%
the evidence will support  a goal ascription which will in turn support the ascription of a belief (say) to the agent of the action.
Accordingly, even those who deny that goal ascription is mindreading should agree that sometimes evidence for goal ascription is indirectly evidence for ascription of belief and other mental states.

\section{slide: Evidence for goal ascription}
What evidence could support ascriptions of goals to actions?

Some philosophers seem to assume that goal ascription depends on knowledge of the agents' intentions
which in turn depends 
on knowledge of their beliefs, desires and other mental states.
If this claim were true,
it would make no sense to discuss the evidential basis of goal ascription except in discussing the evidential basis of mindreading more generally.

However,
the falsity of this claim  is presupposed  in both developmental and comparative research on goal ascription.%
\footnote{ 
Compare \citet{Gergely:1995sq},
	\citet{Woodward:1998dm} and
	\citet{Penn:2007ey}
among many others.
}

To show that goal ascription is possible without ascribing mental states, we need to say what evidence someone could have that would enable her to know to which goals an action was directed.

According to 
what Csibra and Gergely call `the principle of rational action',
%
\begin{quote}
`an action can be explained by a goal state if, and only if, it is seen as the most justifiable action towards that goal state that is available within the constraints of reality.'%
\footnote{
\citet[p.\ 255]{Csibra:1998cx}; cf.\ \citet{Csibra:2003jv}.
A related but different `principle of efficiency' has been formulated by \citet[p.\ 1061]{Southgate:2008el}:
`goal attribution requires that agents expend the least possible amount of energy within their motor constraints to achieve a certain end.'
}
\end{quote}
%
This principle plays two distinct roles.
One role is mechanistic: this principle forms part of an account of how infants (and others) actually ascribe goals.
Another role is normative: this principle also identifies grounds on which it would be rational to ascribe a goal.
I'm concerned only with its normative role.

As Csibra and Gergely formulate it, the principle might seem simple.
But actually their eloquence is hiding some complexity.
It is perhaps worth spelling out what might be involved in applying this principle.
Let me try to spell it out as an inference with premises and a conclusion ...
%
\begin{enumerate}
%
\item action $a$ is directed to some goal;
%
\item actions of $a$'s type are normally capable of being means of realising outcomes of $G$'s type in situations with the salient (to any concerned) features of this situation;
% 
\item no alternative type of action is both 
typically available to agents of this type 
and also 
such that actions of this type would be normally be significantly better\footnotemark \ means of realising outcome $G$ in situations with the salient features of this situation;
\footnotetext{
An action of type $a'$ is a better means of realising outcome $G$ in a given situation than an action of type $a$ if, for instance, actions of type $a'$ normally involve less effort than actions of type $a$ 
in situations with the salient features of this situation 
and everything else is equal; 
or if, for example, actions of type $a'$ are normally more likely to realise outcome $G$ than actions of type $a$
in situations with the salient features of this situation 
and everything else is equal.
}
%
\item the occurrence of outcome $G$ is typically desirable for agents of this type;
%
\end{enumerate}
%
and
%
\begin{enumerate}[resume]
\item there is no other outcome, $G'$, 
the occurrence of which would be at least comparably desirable for agents of this type 
and where (2) and (3) both hold of $G'$ and $a$
%
%where
%	the occurrence of $G'$ would be at least comparably desirable for agents of this type 
%%where
%%	$G$'s occurrence is not a means to $G'$'s nor conversely,
%and
%where  
%	(2) and (3) both hold of $G'$ and $a$
\end{enumerate}
%
may jointly constitute defeasible evidence for the conclusion that:
%
\begin{enumerate}[resume]
\item $G$ is a goal to which action $a$ is directed.
\end{enumerate}
%
We suggest that the above inference, from (1)-(5) to (6), is a route to knowledge of the goals of actions in this sense:
in some cases
it would be possible to know the premises without already knowing the conclusion;
and, 
in some of those cases,
knowing the premises could put one in a position to know the conclusion.%
\footnote{
Knowledge of the conclusion may not require explicit knowledge of 
	what is salient to others,
	what is desirable to them,
	or what makes actions better for them.
It is arguably sufficient that
the individual ascribing a goal 
is entitled to 
rely on being sufficiently similar to the target of her ascription 
with respect to 
these things.
%	what makes one action a better or worse means of realising an outcome than another action for her,  
%	what is desirable to her,
%	and what is salient to her.
}

[\textit{skip paragraph}]
Why accept this?
Suppose that 
	(1)-(5) are facts 
and that
	$G$ is not a goal to which action $a$ is directed.
Then from (1) we know that $a$ has a goal, call it $G'$.
And from (5) we can infer that either 
(i) 
the action, $a$, was not of a type that is normally capable 
of being a means to the outcome to which it was directed,
or
(ii) 
a significantly better (for instance, a more reliable or less effortful) means of achieving $G'$ is typically available to agents of this type,
or
(iii)
$G'$'s occurrence would be significantly less desirable than $G$'s occurrence for agents of this type.
So the agent's action fell short of being appropriate or optimal in some way,
or the agent has atypical preferences or capabilities,
or the situation is not normal.
While any of these is possible---%
and perhaps in some cases even to be expected---%
there are also circumstances in which it is reasonable to suppose that none obtains,
that the situation will be normal in relevant respects,
that agents will have typical preferences and capabilities 
and that they will act in ways that are appropriate and (in the limited sense in play here) optimal.
In such circumstances, the above inference can serve as a route to knowledge of the goals of actions.
The existence of this route to knowledge shows that goal ascription doesn't invariably depend on mental state ascription.

Of course the inference is not deductive.  It is possible for the premises to be true and the conclusion false.  But this is typically of any inference describing a route to knowledge.

I'm not suggesting, of course, that this inference describes how anyone actually ascribes a goal.
My concern is with a related, but more narrowly theoretical question: what evidence could support goal ascription.

I also want to concede that this can't be the whole story about goal ascription  
because this inference has only limited applications.
For example, it cannot be used when an action is the best available means of achieving two or more comparably desirable outcomes.

Anyway, here is one route to knowledge of the goals of actions.
My aim in what follows is to show that there is another route 
to knowledge of the goals of actions 
that is available only to interacting mindreaders
and not to those who merely observe.
But first let me explain why we might want another route to knowledge of the goals of actions.

To apply Csibra and Gergely's principle in a particular case where you first observe an action and then conclude that its goal is $G$,
you need to know that the action is a `justifiable action towards' $G$.
In other words, you need to know that actions of this type are a means to achieving $G$.

Whatever you think of the details of this principle of rational action, 
it seems plausible that the evidence for goal ascription sometimes includes considerations about which ends actions are means to.

\section{The problem of opaque means}
\label{sec:opaque_means}
This the problem of opaque means:
When an observer faces an action but cannot identify ends to which it could be a means, she might be unable to recognise the action's goal.

Ignorance about to which ends actions are means can be an obstacle to goal ascription.

Of course the opaque means are not always  an insurmountable obstacle to goal ascription, and
they may only rarely be a problem for human adults who can rely on culture, convention, or communication.
Where goal ascribers lack such sophistication, the problem of opaque means is likely to be more pressing.

Some of the most plausibly unique aspects of human cognition depend on our abilities to recognise the goals of novel behaviours involving tools, and of communicative gestures.
The problem of opaque means is likely to arise in both cases
if goal ascription is based  entirely on observation 
(so that the possibility of interaction is ignored)
and if the goal ascriber lacks sophisticated social skills.

To illustrate, contrast two cases of tool use.
In one case, someone uses a reamer to  juice  a lime; in the other, someone else scores shag with a lame to prevent a loaf from cracking.
Without communication, repetition or convention,
an observer familiar with reamers but not lames 
may be able to identify the goal of the first action only.

The problem of opaque means is also likely to arise where actions involve multiple steps that do not form a familiar sequence, can occur in various orders and can be interspersed among other activities;
as in preparing spirit from grain, for example.
%\footnote{
%*cut because 'plausibly unique aspects of human cognition'*
%one example is food preparation in some groups of mountain gorillas
%\citep[p.\ 531]{Byrne:2003wx}.
%*This is slightly tricky because Byrne wants a pure behaviour reading account with no understanding of intention.  
%Stress that this is consistent with supposing that goals are ascribed.
%}

The problem of opaque means also affects communicative actions 
 because these characteristically have  goals which the actions are means to realising only because others recognise them as means to realising those goals (a Gricean circle). 
To illustrate, consider an experiment from 
\citet[][experiment 3]{hare_chimpanzees_2004}
 whose two main conditions are depicted in figure \vref{fig:reach_point}.
The pictures in the figure stand for what participants, who were chimpanzees, saw.
The question was whether participants would be able to work out which of two containers concealed a reward.
In the condition depicted in the left panel, participants saw a chimpanzee trying but failing to reach for the correct container. 
Participants had no problem getting the reward in this case, suggesting that they understood the goal of the failed reach.
In the condition depicted in the right panel, a human pointed at the correct container.
Participants did not reliably  get the reward in this case, suggesting that they failed to understand the goal of the pointing action.%
\footnote{
The contrast between the two conditions is not due merely to the fact that one involves a human and the other a chimpanzee.
Participants were also successful when the failed reach was executed by a human rather than another chimpanzee \citep[][experiment 1]{hare_chimpanzees_2004}. 
}
This may be because of the problem of opaque means.
One theoretically possible explanation of these findings is that the participants could identify to which end a failed reach might be a means, but not to which end a communicative gesture might be a means.%
\footnote{\label{fn:communicative_intention}%
Hare and Tomasello 
consider several explanations for their findings including `the hypothesis that chimpanzees do not understand the communicative intent of a cooperative-communicative experimenter' (\citeyear[p.\ 580]{hare_chimpanzees_2004}).
\citet[pp.\ 5--7]{Moll:2007gu} argue for a hypothesis along these lines by appeal to a range of related findings.
}
Whatever the truth about the chimpanzees' performance,
this possibility illustrates how the problem of opaque means can be an obstacle to exploiting communicative gestures.
